Appendix 2: simulating species interactions

### 2015-10-16 Harris_2015/dynamic_simulation.Rmd
new simulations

---
output: pdf_document
title: Appendix 2: simulating species interactions
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

@@

In the large population limit, the number of individuals of species $i$, $n_i$,
evolves over time based on the following differential equation, based on 
multi-species Lotka-Volterra dynamics.  The basic multi-species model is

@@

where $\gamma_{ij}$ is the strength of the mutualism between species $i$ and
species $j$.  The nonlinearity ensures that mutualisms can never eliminate more
than half of the effects of intraspecific competition and destabilize the system
too strongly. In these simulations, each species pair acted either as competitors
$\alpha_{ij} > 0$ or as mutualists $\gamma_{ij} > 0$, but not both. Note that
$\alpha_{ii}$ and $\gamma_{ii}$ were always zero (i.e. intraspecific competition
is modeled outside of the $\alpha$ term and species cannot be their own
mutualists).

@@

```{r, eval=FALSE}
for(n_sites in c(25, 200, 1600)){
  message(n_sites)

  for(rep in 1:n_reps){
    id = paste(n_spp, n_sites, rep, sep = "-")

@@

    saveRDS(results, file = paste0("fakedata/", id, ".rds"))

@@

    # It also expects the data matrices to be transposed
    # relative to the other methods
    write.table(
      t(x_subset), 
      file = paste0("fakedata/", id, "-transposed.txt"), 
      quote = FALSE
    )
  }
}
```

>>>>

---
output: pdf_document
title: Appendix 2: simulating population dynamics with species interactions
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

@@

According to the deterministic skeleton of this model, the number of 
individuals of species $i$, $n_i$, evolves over time based on the following
differential equation, based on multi-species Lotka-Volterra dynamics.  The
basic multi-species model is

@@

where $\gamma_{ij}$ is the strength of the mutualism between species $i$ and
species $j$.  The nonlinearity ensures that mutualisms can never eliminate more
than half of the effects of intraspecific competition and destabilize the system
too strongly. In these simulations, each species pair acted either as competitors
$\alpha_{ij} > 0$ or as mutualists $\gamma_{ij} > 0$, but not both. Note that
$\alpha_{ii}$ and $\gamma_{ii}$ were always zero (i.e. *intra*specific competition
is modeled outside of the $\alpha$ term and species cannot be their own
mutualists).

@@

```{r, eval=TRUE}
for(n_sites in c(25, 200, 1600)){
  message(n_sites)

  for(rep in 1:n_reps){
    id = paste(n_spp, n_sites, paste0("pop", rep), sep = "-")

@@

    write.csv(
      results$observed, 
      file = paste0("fakedata/matrices/", id, ".csv")
    )

    write(
      results$truth, 
      file = paste0("fakedata/truths/", id, ".txt"), 
      ncolumns = length(results$truth)
    )

@@

    # It also expects the data matrices to be transposed
    # relative to the other methods
    write.table(
      t(x_subset), 
      file = paste0("fakedata/matrices/", id, "-transposed.txt"), 
      quote = FALSE
    )
  }
}
```

### 2015-10-16 Harris_2015/simulation.Rmd
new simulations

```{r}
library(progress)
set.seed(1)

thin = 5E3
n_spp = 20
n_reps = 50
```

$$r_e + r_i n_i (1 - \frac{n_i f_i(\vec{n}) + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.

When no propagules enter from outside the system ($r_e=0$) and mutualisms do
not occur ($f_i(\vec{n})$ = 1), this reduces to a multi-species Lotka-Volterra
system

$$r_i n_i (1 - \frac{n_i + \sum_{i \neq j}{\alpha_{ij}n_i n_j}}{K})$$.

where $r_i$ is the [[intrinsic rate of increase]] of species $i$, $n_i$ is its
abundance, and $\alpha_{ij}$ describes the strength of its competition with
species $j$.  In the more general form above, the function $f_i(\vec{n})$ 
acts to reduce the strength of intraspecific competition in the presence of
mutualists.  For these analyses, I used

(1 + exp(0.2 * x / K / n_spp)) / 2

$$f(\vec{n} = $$

[[demographic events happen with probability proportional to corresponding birth/death rate]].  Because these events' probabilities only depend on the 
current state of the system, they form a Markov chain that explores the 
space of possible species assemblages under the population dynamic model.  For
each set of competition and mutualism coefficients, I collected one sample 
after every `r thin`th demographic event.  For the subsequent analysis, 
each species was considered "present" in a sample if its abundance was one
or higher, and "absent" if its abundance in that sample was zero.

```{r}
simulate_communities = function(n_spp = 20,
         K = n_spp / 3,
         maxit = 5E6,
         thin = 1E4,
         seed_rain = rep(.1, n_spp),
         growth = rep(1, n_spp),
         interaction_strength = 1/2,
         prob_negative = 0.75,
         mutualism_maximum = 0.5
){

  # Intraspecific competition can be mitigated by mutualists, up to some
  # maximal amount.  This function rescales the mutualism effect so that
  # it doesn't turn off intraspecific competition entirely.
  mutualism_scaler = function(x, mutualism_maximum){
    (1 + exp(-4 * x / K / n_spp)) * (1 - mutualism_maximum)
  }

  # Randomly generate a vector of interaction strengths (one for each 
  # pair of species)
  interaction_vector = rexp(choose(n_spp, 2), interaction_strength)

  # Randomly mark interactions as positive or negative
  interaction_signs = ifelse(
    rbinom(choose(n_spp, 2), size = 1, prob = prob_negative),
    -1, 
    1
  )

  # Negative interactions are competition; positive ones are mutualisms
  competition_vector = interaction_vector * (interaction_signs == -1)
  mutualism_vector = interaction_vector * (interaction_signs == +1)

  # Fill in the competition and mutualism matrices:
  # Start with a matrix of zeros
  competition = matrix(0, nrow = n_spp, ncol = n_spp)
  # Fill in the upper triangle with the appropriate interaction vector
  competition[upper.tri(competition)] = competition_vector
  # Symmetrize the matrix by adding it to its transpose
  competition = competition + t(competition)

  # Note that intraspecific competition (which would go along the diagonal)
  # is handled separately, and the diagonal is left with zeros.

  mutualism = matrix(0, nrow = n_spp, ncol = n_spp)
  mutualism[upper.tri(mutualism)] = mutualism_vector
  mutualism = mutualism + t(mutualism)

  # Create an empty matrix for storing population sizes
  pops = matrix(nrow = n_spp, ncol = maxit / thin)

  # Initialize the population with a fixed number of individuals per species,
  # depending on the number of species, carrying capacity, and strength of
  # interspecific competition.
  population = rep(ceiling(K / n_spp / mean(competition)), n_spp)

  for(i in 1:maxit){

    # The local growth rate is proportional to population size
    local_growth = growth * population

    # Mutualists act by mitigating intraspecific competition
    mutualist_mitigation = mutualism_scaler(
      population %*% mutualism, 
      mutualism_maximum
    )

    # Add up the two types of interactions affecting death rates
    interactions = population * mutualist_mitigation + population %*% competition

    # Germinating seeds can come from local growth or from outside the system
    birthrates = seed_rain + local_growth

    # Death rates depend on population size, interactions, and carrying capacity
    deathrates = local_growth * interactions / K

    # Generate a vector of vital rates
    rates = c(birthrates, deathrates)

    # Randomly select a birth/death event to occur. Probability
    # of occurring is proportional to the corresponding rate
    event = sample.int(length(rates), 1, prob = rates)

    if(event <= length(birthrates)){
      # Event from first batch, i.e. birth.  Increase the population by one
      index = event
      population[index] = population[index] + 1
    }else{
      # Event from the second batch, i.e. death. If population is nonzero,
      # decrease it by one for the corresponding species
      index = event - n_spp
      if(population[index] > 0){
        population[index] = population[index] - 1
      }
    }

    if(i%%thin == 0){
      # Save the results in the `pops` matrix
      pops[ , i %/% thin] = population
    }
  }

  # Return the "observed" presence-absence vector
  # along with the"true" interaction signs and strengths.
  # The output is transposed because all fitting methods but Pairs
  # require it in that format.
  # Multiplying by one turns TRUEs into 1s and FALSEs into 0s
  list(
    observed = t(pops > 0) * 1,
    truth = interaction_vector * interaction_signs
  )
}
```

```{r, eval=FALSE}
for(n_sites in c(25, 200, 1600)){
  message(n_sites)

  pb = progress_bar$new(total = n_reps)
  for(rep in 1:n_reps){
    id = paste(n_spp, n_sites, rep, sep = "-")

    results = simulate_communities(
      maxit = n_sites * thin, 
      thin = thin, 
      n_spp = n_spp
    )

    pb$tick()

    saveRDS(results, file = paste0("fakedata/", id, ".rds"))

    # Gotelli and Ulrich's Pairs software needs a different format.
    x = results$observed

    # It rejects empty sites, so I remove them here
    x_subset = x[rowSums(x) != 0, colSums(x) != 0]

    # It also expects the data matrices to be transposed
    # relative to the other methods
    write.table(
      t(x_subset), 
      file = paste0("fakedata/", id, "-transposed.txt"), 
      quote = FALSE
    )
  }
}
```

>>>>

# Analyses for Figure 3

## Simulating landscapes from known coefficients

The following function draws random coefficients for a Markov network of a pre-specified size.
The probability distribution was chosen so that most species pairs interact weakly, a few
act strongly, and that most of the interactions were negative rather than positive.

```{r, cache=TRUE}
make_coefficients = function(n_spp){
  # Exponential distribution has lots of mass near 0 but has
  # a long tail.
  true_beta_magnitudes = rexp(choose(n_spp, 2))

  # Species interactions are negative 3/4 of the time in 
  # these simulations.
  b = true_beta_magnitudes * sample(
    c(-1, 1), 
    size = length(true_beta_magnitudes), 
    prob = c(.75, .25),
    replace = TRUE
  )

  # Species' intercepts are normally distributed
  a = rnorm(n_spp, -1)

  # Return the simulated values.
  # The rosalia function stores pairwise parameters in the upper
  # triangle of an n-by-n matrix and stores the species intercepts
  # along the diagonal, so these values are named accordingly.
  c(alpha  = a, beta = b)
}
```

I used the following function to generate "true" parameters with the method above, 
simulate a presence-absence landscape based on those parameters using Markov chain Monte Carlo, and save the results to a "fakedata" folder.

I used Gibbs sampling as my Markov chain Monte Carlo transition operator.
In each round of Gibbs sampling, I cycled through all the species, randomly
updating each one's presence/absence vector in response to its conditional
occurrence probability:

$$p(y_i) = \mathrm{logistic}(\alpha_i + \sum_j\beta_{ij}y_j),$$

where the logistic function is $\frac{1}{1+e^{-x}}$.

```{r, cache = TRUE}
simulate_data = function(n_spp, n_sites, rep_name, n_gibbs = 1000, n_env = 2, sd = 2){

  # Determine the "true" parameters for the simulated assemblage
  par = make_coefficients(n_spp)

  # "True" interaction strengths, to save for later
  truth = par[-(1:n_spp)]

  # "True" intercepts, possibly adjusted below by environment
  alpha = par[1:n_spp]

  # Turn the interaction values into an n-by-n matrix
  # Start with empty matrix; fill in upper triangle; 
  # then fill in lower triangle with its transpose
  beta = matrix(0, n_spp, n_spp)
  beta[upper.tri(beta)] = truth
  beta = beta + t(beta)

  # Environmental states are normally distributed
  env = matrix(rnorm(n_sites * n_env), ncol = n_env)

  alpha_env = matrix(rnorm(n_spp * n_env, sd = sd), nrow = n_env)

  # Simulate the landscape from known process with Gibbs sampling

  # Landscape starts as if betas were all zero. Each species' occurrence probability
  # depends on its alpha value and on the environment (assuming alpha_env is not zero).
  x = matrix(
    plogis(rep(1, n_sites) %*% t(alpha) + env %*% alpha_env), 
    nrow = n_sites, 
    ncol = n_spp
  )

  # Gibbs sampling
  for(i in 1:n_gibbs){

    # Each round of Gibbs sampling updates one species (column) across all sites
    # according to its conditional probability (i.e. conditional on environment
    # and the other species that are present).
    for(j in 1:n_spp){
      x[,j] = rbinom(
        nrow(x),
        size = 1,
        prob = plogis(x %*% beta[ , j] + alpha[j] + env %*% alpha_env[,j])
      )
    }
  }

  # Save the results in a "fake data" folder

  file_stem = paste(n_spp, n_sites, rep_name, sep = "-")

  # Save the matrix of presence/absence observations
  write.csv(
    x, 
    file = paste0("fakedata/matrices/", file_stem, ".csv")
  )

  # Gotelli and Ulrich's Pairs software rejects empty sites, so I remove them here
  x_subset = x[rowSums(x) != 0, colSums(x) != 0]

  # Gotelli and Ulrich's Pairs method expects the data matrices to be transposed,
  # So I save them separately
  write.table(
    t(x_subset), 
    file = paste0("fakedata/matrices/", file_stem, "-transposed.txt"), 
    quote = FALSE
  )

  # Save the "true" species interactions
  write(
    truth, 
    file = paste0("fakedata/truths/", file_stem, ".txt"), 
    ncolumns = length(truth)
  )
}
```

```{r}
set.seed(1)
n_spp = 20

for(n_sites in c(25, 200, 1600)){
  for(type in c("env", "no_env")){
    for(i in 1:50){
      simulate_data(
        n_spp = n_spp, 
        n_sites = n_sites, 
        rep_name = paste0(type, i),
        sd = ifelse(type == "env", 2, 0)
      )
    }
  }
}

```
